% !TEX root = ../../../semexp-thesis.tex

\begin{tblr}{
	colspec = {
		Q[l,h]
		X[l,m]
	},
	column{1} =	{ font=\ttfamily },
	row{1} = 	{ font=\bfseries },
	width =	\textwidth,
	hspan =	minimal,
	rowsep =	1pt,
	colsep =	3pt,
}
	\hline
	%
	Function	& Description \\

	\hline
	%
	{ eval(expression) \\ \hspace*{1em}\textrm{\textit{Example: }}eval("self customer") }	& Evaluate a Smalltalk expression in the context of the explored object and return the result or error. Can be executed in isolation. \\
	evalAndReturn(expression)	& Evaluate a Smalltalk expression in the context of the explored object and pass back the result to the sender of the original semantic message. Only available if the agent was invoked through a semantic message. \\
	browsePackage(packageName)	& Return a hierarchical list of classes within a package. \\
	browseClass(className)	& Enumerate all methods defined on a class or one of its superclasses or their metaclasses (for static methods), grouped by the defining class and the method category (protocol) within the class organization. \\
	browseMethod(className, selector)	& Retrieve the source code of a method defined in a class. \\
	{ browseSenders(selector[, query]) \\ \hspace*{1em}\textrm{\textit{Examples:}}\\\hspace*{1em}browseSender("printOn:")\\\hspace*{1em}browseSender("printOn:",\\\hspace*{2em}\llap{\raisebox{0.2ex}[0ex][0ex]{\ensuremath{\rcurvearrowse\hspace*{2pt}}}}"date yyy-mm-dd") }	& Search all methods in the system that send messages with the name of a selector and return a representative sample of methods. If an optional query argument has been provided, rank the methods based on their relevance to the query. \\
	\hline
\end{tblr}
