% !TEX root = semexp-thesis.tex

\begingroup
\let\raggedsection\centering

\chapter*{Acknowledgments}
\label{cha:acknowledgments}
\endgroup
\begin{quotation}
	\noindent
	While I have started writing this thesis six months ago, much of the work presented here originates from the past one and a half years, or even before that, considering my broader orientation within the exploratory programming field.
	I am grateful to several people who made this research not only an insightful but also a very fulfilling experience for me.

	First and foremost, to Marcel Taeumel for mentoring and advising me throughout this research as well as several others before. Thank you for collaborating on multiple papers; for sharing your experience and ideas about exploratory programming, academia, and Squeak during many extensive and delightful discussions; for providing me with \emph{significant} feedback on my writing, always lending me an ear or eye when I bombarded you with too many questions, and critically shaping my entire master's studies into something wonderful.
	To Lukas Böhme and Toni Mattis for contributing their ideas and perspectives of machine learning to my research, and again to Lukas for reviewing my writing.
	To Robert Hirschfeld for motivating me to study this field, making this project possible (including financial support), and helping me on various other matters.

	I am also thankful to other members and students of the SWA group for discussing my ideas; to the three anonymous reviewers of the Onward! paper for providing their inspiring and detailed reviews; and to u/plasticpears for discussing their vision of AI-augmented exploratory programming systems with me\footnote{\url{https://www.reddit.com/r/smalltalk/comments/1b3dx4q/smalltalk_llms}}.
	I also thank Patrick Rein and Willy Scheibel for the opportunity of working together on earlier projects: Patrick for showing me the first steps of educational writing and Willy for sharing an elaborate mindset of academic thinking and writing with me.

	This thesis would not look like this without the ecosystem and the individual support by Tobias Pape and the \TeX.SE community.

	Finally, I would like to thank my family for reading my thesis and for all your other invaluable support.
\end{quotation}
